---
title: "ANOVA_Examples"
output:
  html_document:
    df_print: paged
---

# General Notes

1. Fixed effects ANOVA
  + $H_o$ is testing whether means for all groups are equal.
  + F-statistic uses MSE for denominator (MSE = Expected Mean Squares)
2. Random effects ANOVA 
  + $H_o$ testing if treatment variances = 0.
  + F-statistic uses Expected Mean Squares (i.e., Mean Square Within) for denominator
  + Each source of variation will consist of a linear combination of $\sigma_e^2$ plus variance components whose subscript matches at least one letter in the source of variation.
  + The coefficients for the identified variance components will be the letters not found in the
subscript of the variance components.
3. Mixed Models
  + If Factor B is nested within Factor A, there can be no interaction.
  + Subsampling is treated as a random effect nested within treatments.

# R-isms
1. aov() can only be used for balanced designs
2. lmer() can only be used for models with random effects
  + If we are doing a LR test to compare models, make sure to set REML = FALSE in lmer()

```{r setup, include=FALSE}
rm(list=ls())

options("scipen" = 1, "digits" = 4)

library(car) # Anova function for Type II or Type III Anova
library(emmeans)  # emmeans function (lsmeans deprecated)
library(lme4) # lmer
```

# List of Examples
1. [Two-Way Anova, Balanced, r = 1]
  + [Fixed Effects]
    - [No interaction term]
    - [Nested term - saturated model]
    - [With interaction - saturated model]
  + [Mixed Models]
    - [Stim-Fixed, Plot-Random, No interaction term]
2. [Two-way ANOVA Balanced, r = 2]
  + [Fixed Effects with interaction term]
  + [Fixed effect, nested]
  + [Mixed model, nested]

3. [Two-Way ANOVA Unbalanced, Subsampling]

# Two-Way Anova, Balanced, r = 1
```{r}
stim = as.factor(c(rep("S1",4), rep("S2", 4), rep("S3", 4), rep("S4", 4)))
plot = as.factor(rep(c("1", "2", "3", "4"), 4))
rootwt = 
c(3.4, 3.2, 3.233, 3.067,
  3.833, 3.733, 3.6, 3.55,
  3.9, 3.7, 3.35, 3.65,
  4.33, 3.933, 4.067, 3.867)

interaction.plot(plot, stim, rootwt)

```

## Fixed Effects

### No interaction term

$$Y_{ij} = \alpha_i + \beta_j + \epsilon_{ij}$$

 Source of Variation | df  | EMS | F  
-------------------- | --- | --- | --- 
A (Stim)            | $4-1 = 3$ | $\sigma_e^2 + b \frac{\sum_i\alpha_i^2}{a}$ | $\frac{EMS_A}{MS_E}$
B (Plot) | $4-1 = 3$| $\sigma_e^2 + a \frac{\sum_j \beta_j^2}{b}$ | $\frac{EMS_B}{MS_E}$
Error | $9$ |  $\sigma_e^2$
Total | $16-1 = 15$      |  


```{r}
model.aov = aov(rootwt ~ plot + stim)
summary(model.aov)
```

### Nested term - saturated model

$$Y_{ij} = \alpha_i + \beta_{j(i)} + \epsilon_{ij}$$

 Source of Variation | df | EMS | F  
--- | --- | --- | --- 
A (Stim)          | $4-1 = 3$ | $b\frac{\sum_i\alpha_i^2}{a}$ 
B (Plot) (Plot(Stim)) | $4(4-1) = 12$ | $\frac{\sum_i\sum_j\beta_{j(i)}^2}{ab}$
Total | $16-1 = 15$     |  

```{r}
model.aov = aov(rootwt ~ stim/plot)
#model.aov = aov(rootwt ~ stim + stim:plot)   # Equivalent model
summary(model.aov)
```

### With interaction - saturated model

$$Y_{ij} = \alpha_i + \beta_j + (\alpha\beta)_{ij} + \epsilon_{ij}$$

 Source of Variation | df | EMS |   
--- | --- | --- | --- 
A (Stim) | $4-1 = 3$ | $b\frac{\sum_i\alpha_i}{a}$ 
B (Plot) | 4 -1 = 3 | $a\frac{\sum_j\beta_j}{b}$ 
A:B (Stim:Plot) | (4 -1)(4-1) = 9 | $\frac{\sum_i\sum_j(\alpha\beta)_{ij}}{ab}$  
Total | $16-1 = 15$     |  

```{r}
model.aov = aov(rootwt ~ plot + stim + plot:stim)
#model.aov = aov(rootwt ~ plot * stim)   # Equivalent model
summary(model.aov)
```

## Mixed Models

### Stim-Fixed, Plot-Random, No interaction term

$$Y_{ij} = \alpha_i + \beta_j + \epsilon_{ij}$$

 Source of Variation | df | EMS  | F
--- | --- | --- | --- 
A (Stim) | $4-1 = 3$ | $\sigma_e^2 + b\frac{\sum_i \alpha_i^2}{a}$ | $\frac{EMS_A}{MS_E}$
B (Plot) | $4-1 = 3$     | $\sigma_e^2 + a\sigma_b^2$ | $\frac{EMS_B}{MS_E}$
Error | $(4-1)(4-1) = 9$ |  $\sigma_e^2$
Total | 16-1 = 15 |  

```{r}
# stim - fixed effect
# plot - random effect
model.lmer = lmer(rootwt ~ 1 + stim + (1|plot))
summary(model.lmer)
anova(model.lmer)
```

```{r}
model.aov = aov(rootwt ~ stim + Error(plot))
summary(model.aov)
```

# Two-way ANOVA Balanced, r = 2
```{r}
stim = as.factor(c(rep("S1",8), rep("S2", 8), rep("S3", 8), rep("S4", 8)))
plot = as.factor(rep(c("1", "1", "2", "2", "3", "3", "4", "4"), 4))
plot_unique = as.factor(c("S1-1", "S1-1", "S1-2", "S1-2", "S1-3", "S1-3", 
                          "S1-4", "S1-4",
                          "S2-1", "S2-1", "S2-2", "S2-2", "S2-3", "S2-3", 
                          "S2-4", "S2-4",
                          "S3-1", "S3-1", "S3-2", "S3-2", "S3-3", "S3-3",
                          "S3-4", "S3-4",
                          "S4-1", "S4-1", "S4-2", "S4-2", "S4-3", "S4-3",
                          "S4-4", "S4-4"))
rootwt = 
c(3.3, 3.5, 3.5, 3.0, 3.4, 3.1, 3.3, 2.9,
  3.7, 4.0, 3.5, 4.0, 3.4, 3.8, 3.4, 3.7,
  3.8, 4.0, 3.6, 3.8, 3.3, 3.4, 3.6, 3.7,
  4.3, 4.4, 3.8, 4.1, 3.9, 4.2, 3.7, 4.0)

interaction.plot(plot, stim, rootwt)

```

## Fixed Effects with interaction term

$$Y_{ijk} = \alpha_i + \beta_j + (\alpha\beta)_{ij} + \epsilon_{ijk}$$

where $i = 1...a, j = 1...b, k = 1..r$

 Source of Variation | df | EMS | F  
--- | --- | --- | --- 
A (Stim) | $4-1 = 3$ | $\sigma_e^2 + rb\frac{\sum_i\alpha_i}{a}$ | $\frac{EMS_A}{MS_E}$
B (Plot) | $4 -1 = 3$ | $\sigma_e^2 + ra\frac{\sum_j\beta_j}{b}$ | $\frac{EMS_B}{MS_E}$
A:B (Stim:Plot) | $(4 -1)(4-1) = 9$ | $\sigma_e^2 + r\frac{\sum_i\sum_j(\alpha\beta)_{ij}}{ab}$ | \frac{EMS_AB}{MS_E}
Error | $16$ | $\sigma_e^2$
Total | $32-1 = 31$     |  

```{r}
model.aov = aov(rootwt ~ plot + stim + plot:stim)
#model.aov = aov(rootwt ~ plot * stim)   # Equivalent model
summary(model.aov)
```

# Fixed effect, nested

$$Y_{ijk} = \alpha_i + \beta_{j(i)} + \epsilon_{ijk}$$

where $i = 1...a, j = 1...b, k = 1..r$

 Source of Variation | df | EMS | F  
--- | --- | --- | --- 
A (Stim) | $4-1 = 3$ | $\sigma_e^2 + rb\frac{\sum_i\alpha_i}{a}$ | $\frac{EMS_A}{MS_E}$
B (Plot) | $4(4-1) = 12$ | $\sigma_e^2 + r\frac{\sum_j\beta_{j(i)}}{ab}$ | $\frac{EMS_B}{MS_E}$
Error | $16$ | $\sigma_e^2$
Total | $32-1 = 31$     |  

```{r}
model.aov = aov(rootwt ~ stim/plot)
# model.aov = aov(rootwt ~ stim + stim:plot)
summary(model.aov)
```


## Mixed model, nested

$$Y_{ijk} = \alpha_i + \beta_{j(i)} + \epsilon_{ijk}$$

where $i = 1...a, j = 1...b, k = 1..r$

 Source of Variation | df | EMS | F  
--- | --- | --- | --- 
A (Stim) | $4-1 = 3$ | $\sigma_e^2 + r\sigma_{b(a)}^2 + rb\frac{\sum_i\alpha_i}{a}$ | $\frac{EMS_A}{MS_{B(A)}}$
B(A) (Plot(Stim)) | $4(4 -1) = 12$ | $\sigma_e^2 + r\sigma_{b(a)}^2$ | $\frac{EMS_B}{MS_E}$
Error | $16$ | $\sigma_e^2$
Total | $32-1 = 31$     |  


```{r}
# stim - fixed effect
# plot - random effect, plot nested in stim
model.lmer = lmer(rootwt ~ 1 + stim + (1|stim:plot))
# The following model is equivalent
# model.lmer = lmer(rootwt ~ 1 + stim + (1|plot:stim))
summary(model.lmer)
anova(model.lmer)
```


```{r}
# Check replications
df = data.frame(rootwt = rootwt, stim = stim, plot = plot)
replications(rootwt ~ stim + Error(plot:stim), data = df)
model.aov = aov(rootwt ~ stim + Error(stim:plot))
summary(model.aov)
```

```{r}
# Alternate method to handle nesting
summary(aov(rootwt ~ stim + Error(plot_unique)))
```

## Two-Way ANOVA Unbalanced, Subsampling

```{r}
# Use lmer function
stim = as.factor(c(rep("S1",15),rep("S2",16),rep("S3",14),rep("S4",12)))
plotS1 = as.factor(c(rep("1",3),rep("2",3),rep("3",3),rep("4",3),rep("5",3)))
plotS2 = as.factor(c(rep("1",3),rep("2",3),rep("3",3),rep("4",2),rep("5",3),rep("6",2)))
plotS3 = as.factor(c(rep("1",3),rep("2",3),rep("3",2),rep("4",2),rep("5",2),rep("6",2)))
plotS4 = as.factor(c(rep("1",3),rep("2",3),rep("3",3),rep("4",3)))
plot = as.factor(c(plotS1,plotS2,plotS3,plotS4))
rootwt = 
c(3.3,3.4,3.5,3.1,3.5,3.0,3.2,3.1,3.4,3.3,2.9,3.0,3.3,3.3,3.1,  
  3.8,3.7,4.0,3.5,3.8,3.9,3.6,3.4,3.8,3.4,3.7,3.6,3.7,3.6,3.5,3.9, 
  3.8,3.9,4.0,3.6,3.7,3.8,3.3,3.4,3.6,3.7,3.5,3.9,3.4,3.7,
  4.3,4.3,4.4,4.1,3.9,3.8,4.2,4.1,3.9,3.7,3.9,4.0)
interaction.plot(plot, stim, rootwt)
```

```{r}
# stim - fixed effect
# plot - nested in stim, random effect
# cores - nested within plots within stim, random effect
model.lmer = lmer(rootwt ~ stim + (1|plot:stim))
#summary(model.lmer)
anova(model.lmer)
```

```{r}
emmeans(model.lmer, "stim")
```

