---
title: "Bayes Capture-Recapture Example"
author: "Melissa Wong"
date: "6/16/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(out.width = "50%")
knitr::opts_current$set(fig.align = "center")
library(tidyverse)
library(RColorBrewer)

options(ggplot2.discrete.color=brewer.pal(8, "Dark2"))
options(ggplot2.discrete.fill=brewer.pal(8, "Dark2"))
```


# One-parameter, closed system model

Makes the following assumptions:

* All bugs are equally likely to be found
* Both test teams are equally likely to find bugs
* Findings for each test are independent
* Total number of bugs is fixed

Given these assumptions, then the likelihood for _y_ bugs found by both test teams is given by the hypergeometric distribution

$$p(y_i;N, n_i, k_{i-1}) = \frac{\binom{k_i}{y_i}
                                  \binom{N-k_{i-1}}{n_i-y_i}}{\binom{N}{n_i}}$$

where $y_i$ is the number of previously marked bugs found on recapture, $n_i$ is the sample size of test $i$ and $k_{i-1}$ is the number of previously marked bugs.

We define a prior for $N$, so the model is

\begin{align*}
  p(N | y, n, k) &\propto \prod_i \frac{\binom{k_i}{y_i}\binom{N-k_i}{n_i-y_i}}{\binom{N}{n_i}} * p(N) \\
  p(N) &\propto Uniform(1, 200)
\end{align*}

```{r}
# Simulate data
N <- 50
n <- c(10, 20)
theta <- rep(NA, 2)
theta[1] <- rhyper(1, N, 0, n[1])
theta[2] <- rhyper(1, theta[1], N-theta[1], n[2])
```

```{r}
# Simulate 1st capture and mark
capture <- tibble(N = seq(0:201),
                  prior = c(0, rep(0.005, 200), 0))
lik <- dhyper(theta[1], capture$N, 0, n[1])
lik[is.na(lik)] <- 0
post_unstd <- lik * capture$prior
capture$post1 <- post_unstd / sum(post_unstd)
```


```{r echo=FALSE}
# See https://stats.stackexchange.com/questions/381520/how-can-i-estimate-the-highest-posterior-density-interval-from-a-set-of-x-y-valu

# Find approximate HPDI
hdi = function(x, x.density, coverage)
{
  best = 0
  for (ai in 1 : (length(x) - 1))
  {
    for (bi in (ai + 1) : length(x))
    {
      mass = sum(diff(x[ai : bi]) * x.density[(ai + 1) : bi])
      if (mass >= coverage && mass / (x[bi] - x[ai]) > best)
      {
        best = mass / (x[bi] - x[ai])
        ai.best = ai
        bi.best = bi
      }
    }
  }
  c(x[ai.best], x[bi.best])
}
```

```{r}
# Simulate recapture
# posterior from 1st capture is now prior for recapture
lik <- dhyper(theta[2], n[1], capture$N-n[1], n[2])
lik[is.na(lik)] <- 0
post_unstd <- lik * capture$post1
capture$post2 <- post_unstd / sum(post_unstd)

# Calculate 95% HPDI
hpdi <- capture$N[hdi(capture$N, capture$post2, 0.95)]

capture %>%
  pivot_longer(-N, names_to="type", 
               values_to="prob") %>%
  ggplot() +
  geom_line(aes(x=N, y=prob, color=type)) +
  geom_vline(xintercept=hpdi, 
             linetype="dotted")
```

* Total number of simulated bugs: `r N`.
* Initial mark group size (all tagged): `r n[1]`
* Recapture group size: `r n[2]`
* Recapture group, number previously tagged: `r theta[2]`

The 95% highest posterior density interval is `r hpdi`.

The maximum a posteriori estimate for N is `r capture$N[which.max(capture$post2)]`.

# Three-parameter model, closed system

Makes the following assumptions:

* The test teams are not equally proficient 
* Findings for each test are independent
* Total number of bugs is fixed

Given these assumptions, then the likelihood for _k_ bugs found by both test teams is described by a multinomial distribution

$$p(k|N, p1, p2) = \frac{N!}{k_{00}!k_{01}!k_{10}!k_{11}!}
q_{00}^{k_{00}}q_{01}^{k_{01}}q_{10}^{k_{10}}q_{11}^{k_{11}}$$

where 

\begin{align*}
  k_{10} &= \text{bugs found by first test and not second} \\
  k_{01} &= \text{bugs found by second test and not first} \\
  k_{11} &= \text{bugs found by both tests} \\
  k_{00} &= N - k_{10} - k{01} - k{11} \text{bugs found by neither test} \\
  p1 &= \text{probability test1 finds bug} \\
  p2 &= \text{probability test2 finds bug} \\
  q_{00} &= (1-p1) * (1-p2) \\
  q_{10} &= p1 *  (1-q2) \\
  q_{01} &= (1-p1)*p2\\
  q_{11} &= p1 * p2
\end{align*}

And we need priors for $N, p1, p2$; let's keep same prior for N and Uniform[0.1,1] for $p1, p2$.


```{r}
# Simulate 1st capture and mark
prior_p1 <- tibble(p1 = seq(0, 1, length.out=11),
                   prob1 = c(0, rep(0.1, 10)))

prior_p2 <- tibble(p2 = seq(0, 1, length.out=11),
                   prob2 = c(0, rep(0.1, 10)))

prior_N <- tibble(N = seq(0:201),
                  prob = c(0, rep(0.005, 200), 0))

grid <- crossing(prior_N, prior_p1, prior_p2) %>%
  mutate(lik = 0,
         post_unstd = 0)

k10 <- n[1] - theta[2]
k01 <- n[2] - theta[2]
k11 <- theta[2]

for(i in seq_along(grid$N))
{
  k00 <- grid[i,]$N - k10 - k01 - k11
  if (k00 < 0)
  {
    grid[i,]$lik <- 0.0
    grid[i,]$post_unstd <- 0.0
  }
  else{
    q00 <- (1-grid[i,]$p1) * (1-grid[i,]$p2)
    q10 <- grid[i,]$p1 * (1-grid[i,]$p2)
    q01 <- (1-grid[i,]$p1) * grid[i,]$p2
    q11 <- grid[i,]$p1 * grid[i,]$p2
    grid[i,]$lik <- dmultinom(x=c(k00, k10, k01, k11),
                             prob = c(q00, q10, q01, q11))
    grid[i,]$post_unstd <- grid[i,]$lik * grid[i,]$prob * grid[i,]$prob1 * grid[i,]$prob2
  }
}

grid$post <- grid$post_unstd/sum(grid$post_unstd)
```


```{r}
# Plot N
postN <- grid %>%
  group_by(N) %>%
  summarize(posterior=sum(post)) %>%
  mutate(prior = prior_N$prob)

# Calculate 95% HPDI
hpdi <- postN$N[hdi(postN$N, postN$posterior, 0.95)]

postN %>%
  pivot_longer(-N, names_to="type", values_to = "prob") %>%
ggplot() +
  geom_line(mapping=aes(x=N, y=prob, color=type)) +
  geom_vline(xintercept=hpdi, 
             linetype="dotted")
```

* Total number of simulated bugs: `r N`.
* Initial mark group size (all tagged): `r n[1]`
* Recapture group size: `r n[2]`
* Recapture group, number previously tagged: `r theta[2]`

The 95% highest posterior density interval is `r hpdi`.

The maximum a posteriori estimate for N is `r postN$N[which.max(postN$posterior)]`.

```{r}
postp1 <- grid %>%
  group_by(p1) %>%
  summarize(prob=sum(post))

postp2 <- grid %>%
  group_by(p2) %>%
  summarize(prob=sum(post))
```

