---
title: "Confidence Intervals for Proportions"
author: "Melissa Wong"
date: "4/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(binom)
library(ggplot2)
library(dplyr)
library(tidyr)
```

# Description

This is a quick simulation to compare the effect of three different priors on the Bayesian credible interval coverage.  

# Priors

The three priors considered are plotted below.

```{r}
x <- seq(0.0, 1.0, 0.01)
ggplot() +
  geom_line(mapping=aes(x=x, y=dbeta(x, 0.5, 0.5),
            color="alpha=beta=0.5")) +
  geom_line(mapping=aes(x=x, y=dbeta(x, 1, 1),
            color="alpha=beta=1")) +
    geom_line(mapping=aes(x=x, y=dbeta(x, 2, 2),
            color="alpha=beta=2")) +
  labs(title="Beta priors",
       x="proportion", y="probability density")
```

# Coverage

The coverage for the three Bayesian credible intervals is compared to the Agresti-Coull coverage in the plots below.

```{r}
N <- 100
probs <- tibble(p = seq(0.01, 0.99, 0.01))
sims <- cbind(probs, N, t(apply(probs, 1, function(x) rbinom(1000, N, prob=x)))) %>%
  pivot_longer(cols=-c(p, N), names_to="iter",
                             values_to="x")

# Calculate CIs for the three priors
CIs <- rbind(sims %>% mutate(shape=0.5),
             sims %>% mutate(shape=1),
             sims %>% mutate(shape=2)) %>%
  mutate(res=binom.bayes(x, N, prior.shape1=shape, 
                         prior.shape2=shape)[7:8],
         resAC=binom.agresti.coull(x,N)[5:6]) %>%
  mutate(isCovered = res$lower <= p & p <= res$upper) %>%
  mutate(isCoveredAC = resAC$lower <= p & p <= resAC$upper)

CIs %>%
  group_by(p, shape) %>%
  summarize(coverage = mean(isCovered),
            coverageAC = mean(isCoveredAC)) %>%
  ggplot() +
  geom_line(mapping=aes(x=p, y=coverage,
                        color=factor(shape))) +
  geom_line(mapping=aes(x=p, y=coverageAC), 
            color="grey") +
  facet_wrap(~shape, nrow=3)
```
# Conclusion

The Jeffries prior ($\alpha=\beta=0.5$) has the worst coverage properties.  The $\alpha=\beta=1$ and $\alpha=\beta=2$ priors show better coverage properties. Of note, as the probability $\to 0/1$, the $\alpha=\beta=1$ prior tends to be conservative whereas the $\alpha=\beta=2$ is too optimistic.

