---
title: "Mixed Model Example"
author: "Melissa Wong"
date: "4/27/2020"
output: pdf_document
---

```{r echo=FALSE, results='hide', message=FALSE, warning=FALSE}
rm(list=ls())

options("scipen" = 1, "digits" = 4)

library(tidyverse)
library(corrplot)
```

# Read in Data

```{r}
# Notes about data
# sleep is continuous (actual hours of sleep)
# numpains is continuous (actual number of areas of the body that were hurting as a result #of physical activity on a particular day - knees and ankles being the most common).
# Nutrition, fluids, physical state, mental state, quality of workout and rate of exertion are self-rated items. 
#Illness and sportpain are yes/no. 
#Relaxin phase is a three-level factor.

# make qualitative features factors

athletes_orig <- read.table("./data.txt", header=TRUE) %>%
  mutate_at(vars(PlayerNum, Nutrition, Illness, Physical.State, Mental.State,
                 Relaxin2, WO_typeCode, Quality_WkO, Rate.exertion, SportPain),
            list(factor)) 

summary(athletes_orig)
```

# Check missing data

Looks like there may be rows with multiple missing predictors; let's check if it's all from the same player.

```{r}
athletes_orig %>% filter(is.na(AgeYrs))
```

Looks like player 8 is a problem; missing data for predictors of interest (Relaxin, Age, Team3Grps) as well as responses (SportPain and NumPain)

```{r}
# Drop player 8 
# Drop the rows where Relaxin is NA since that's a predicotor we care about
# Drop the rows where SportPain are NA since that's the response we care about
# Drop columns that won't be used in the model
athletes <- athletes_orig %>%
  filter(!is.na(Relaxin2) & !is.na(NumPains) & !is.na(SportPain)) %>%
  filter(PlayerNum != 8) %>%
  select(-Index, -Player.order,-AgeYrs, -Fluids, -WO_typeCode, -Quality_WkO, 
           -Rate.exertion, -Workout.time) %>%
  mutate(Index = row_number()) %>%
  group_by(PlayerNum) %>%
  mutate(Measurement = row_number()) %>%
  ungroup()

summary(athletes)
anyNA(athletes)
```

Do a quick sanity check that there aren't any rows where NumPains > 0 but SportPain = 0 or
NumPains = 0 and SportPain = 1
```{r}
athletes %>%
  filter((NumPains > 0 & SportPain == 0) | (NumPains == 0 & SportPain == 1))
```

Now let's look at relationships between predictors in the clean data.
```{r}
pairs(athletes)
```

Let's zoom in on BMI/Height/Weight.  There's a strong positive correlation between Weight and both Height so I won't use Weight in the model.

```{r}
pairs(~ BMI + Height_in + Weight_lbs, data=athletes)
```

Let's also zoom in on Age & Team3Grp.  As expected, the Varsity group is generally older than the JV group which is likewise older than the 9th group. I'm just going to use Team3Grps in the model.

```{r}
pairs(~ AgeMonths + Team3Grps, data=athletes)
```

# Create models

To simplify things I'll only use the predictors \textit{BMI, Height_in, Team3Grp, Relaxin2}.  I'll consider two models: 1) logistic regression for P[SportPain = 1] and 2) a GLM for NumPains.

First, check whether or not data is balanced.

```{r}
athletes %>%
  group_by(PlayerNum) %>%
  summarise(count = n())
```

Data is unbalanced, so 