---
title: "Survival Analysis Example"
author: "Melissa Wong"
date: "4/19/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

# Description

Example comparing original method for calculating $P[sync]$ vs using survival analysis.

# Simulate Data

```{r}
N <- 100
# Time until successful sync ~ Normal with mean=2 sec, sd=0.5
df <- tibble(time = rnorm(N, mean = 2, sd = 0.5),
             success = 1)
# Simulate 30% of runs not syncing within 5 sec (max run duration)
ids <- sample(1:N, 30)
df$success[ids] = 0
df$time[ids] = 5
```

# Original Analysis

Calculate probability of successful sync with Beta(1,1) prior so $P[sync] = Beta(\#successes+1, \#failures+1)$

```{r}
library(binom)

nsuccess <- sum(df$success)
(res <- binom.bayes(nsuccess, N, type="highest", prior.shape1 = 1,
            prior.shape2 = 1))
binom.bayes.densityplot(res)
```

And the boxplot of time until successful sync.

```{r}
df %>%
  filter(success == 1) %>%
  ggplot(mapping=aes(x=time, y="Success")) +
  geom_boxplot() +
  geom_jitter()
```

# Improved method

Use survival analysis (specifically Kaplan-Meier) which makes use of the time to event and gives a more informative distribution of P[sync].

In this example, the "event" of interest is successful sync.

Note: Kaplan-Meier only applicable for single categorical predictor.  Use Cox proportional hazard model for multiple/quantitative predictors.

```{r}
library(survival)
library(survminer)

fit <- survfit(Surv(time, success)~1, data=df)
#summary(fit)
```
```{r}
ggsurvplot(fit, conf.int = TRUE, surv.median.line = "hv") +
  labs(y="Probability of No Sync")
```


```{r include=FALSE}
ggsurvplot(fit, conf.int = TRUE, fun="cumhaz") +
  labs(title="Cumulative Hazard")
```
```{r}
ggsurvplot(fit, conf.int = TRUE, surv.median.line = "hv",
           fun="event") +
  labs(y="Probability of Sync")
```
This is a more informative analysis because we don't just get a single overall estimate for $P[sync]$ but rather $P[sync]$ as a function of time.

# Cox Model Example

```{r}
data("lung")
head(lung)

fit.cox <- coxph(Surv(time, status) ~ sex, 
                 data = lung)
summary(fit.cox)
# Baseline
# ggsurvplot(survfit(fit.cox, data=lung), 
#            conf.int = TRUE,
#            surv.median.line = "hv")

sex_df <- with(lung,
               data.frame(sex = c(1, 2)))
fit <- survfit(fit.cox, newdata = sex_df)
ggsurvplot(fit, conf.int = TRUE, data=sex_df, legend.labs=c("Sex=1", "Sex=2"),
           ggtheme = theme_minimal())
```

```{r}
fit.km <- survfit(Surv(time, status) ~ sex, 
                 data = lung)
ggsurvplot(fit.km, conf.int = TRUE, surv.median.line = "hv")
```

