---
title: "Bayes R Packages Part 1"
author: "Melissa"
date: "10/25/2020"
output: 
  github_document:
    toc: true
    toc_depth: 2
---

# Introduction

In Part 1, I skipped over model diagnostics so I'm going to address that in this post. I'll demonstrate _bayesplot_ functions here, but _shinystan_ is a nice, interactive alternative.

# Setup Environment

Recall that we're looking at the following model for the _mtcars_ data set:

$$mpg = a + b*disp + \epsilon$$

```{r results='hide', message=FALSE, warning=FALSE}
rm(list=ls())

library(tidyverse)
library(rstanarm)
library(rethinking)
library(bayesplot)
library(ggplot2)
library(rstan)
#library(tidybayes)

#knitr::opts_chunk$set(out.width = "50%")
knitr::opts_chunk$set(fig.align = "center")
knitr::opts_chunk$set(message=FALSE)
knitr::opts_chunk$set(warning=FALSE)

options("scipen" = 1, "digits" = 4)

library(datasets)
data(mtcars)
```

# Package _rstanarm_

I'll demonstrate first using the model generated from _stan_glm_ with default priors.

```{r results='hide'}
mdl1 <- stan_glm(mpg ~ disp, data = mtcars)
```

Once the model has been fit, we can use either _as.matrix_ or _as.array_ to extract the posterior draws.  The key difference is that _as.array_ keeps the chains separate.

```{r}
post <- as.array(mdl1)
str(post)
```
Note that the default is four chains; we can specify the number of chains with the _chains=_ argument in _stan_glm_.

## Trace Plots with _bayesplot_

The _bayesplot_ package provides the function _mcmc_trace_ which plots the MCMC draws.

```{r}
mcmc_trace(post, pars=c("disp", "sigma"))
```
There are three things we are looking for in the trace plot of each chain:

  1. *Good mixing* -  In other words, the chain is rapidly changing values across the full region versus getting "stuck" near a particular value and slowly changing.
  
  2. *Stationarity* - The mean of the chain is relatively stable.
  
  3. *Convergence* - All of the chains spend most of the time around the same high-probability value.
    
The trace plots above look good.  However, sometimes it can be hard to tell when there are multiple chains overlaid on the same plot, so two alternatives are shown below.

## Trace Rank Plots with _rethinking_

One alternative is to use the _trankplot_ function from the _rethinking_ package.  This function plots a trace rank plot which is the distribution of the ranked samples.

```{r}
# Note that I specify the $stanfit parameter from the stan_glm object
trankplot(mdl1$stanfit, pars=c("disp", "sigma"))
```

## Trace Plots with _ggplot2_

Another alternative is to manually plot each chain separately.  Here's one way to do it with _ggplot2_.

```{r}
library(gridExtra)

pars <- c("disp", "sigma")

plts <- list()
for (par in pars)
{
  df <- as.data.frame(post[,,par]) %>%
    mutate(iteration = row_number()) %>%
    pivot_longer(cols=c(-"iteration"), values_to="value", names_to="chain")

  plts[[par]] <- df %>%
    ggplot() +
    geom_line(mapping=aes(x=iteration, y=value), color="blue") +
    facet_wrap(~chain, ncol=1) +
    labs(title=par)
}

grid.arrange(grobs=plts, nrow=1)
```


